---
title: "Mapping DHS Survey Estimates to ADM Level 2 Geographies"
author: "Corey Sparks PhD"
date: "4/14/2021"
output: html_document
---

In this brief example, I show how to use the GPS coordinates from the [Demographic and Health Survey data]() and merge them to the ADM2 subnational geographic level for the country of Ethopia. Then I produce estimates using the DHS data for ADM 2 regions of the country.

This is possible by useing the GIS capacity of the `sf` package to spatially intersect the DHS points and the ADM 2 polygons. 

```{r}
library(sf)
library(mapview)
```

## Read in dhs points
```{r}
ethpoly <- st_read(dsn = "~/OneDrive - University of Texas at San Antonio//students/fikre/spatial_epi/ETH_adm2.shp")

ethpoly$struct <- 1:dim(ethpoly)[1]

plot(ethpoly["struct"])

```


## Read in dhs sample locations and ADM 2 regions.

The adm2 shapefile can be found in the [Diva GIS international data repository](https://www.diva-gis.org/gdata), or from the [IPUMS International site](https://international.ipums.org/international/gis_harmonized_2nd.shtml) below I use the ADM2 level of administrative geography. 

These locations are not identified in the DHS, but by performing a spatial intersection, we can merge the DHS survey locations to the ADM 2 units



```{r}
eth_dots<-st_read("~/OneDrive - University of Texas at San Antonio//students//fikre/ethiopia_dhs/ETGE52FL/ETGE52FL.shp")
eth_dots <- eth_dots[eth_dots$LATNUM>0,]
```

```{r}
eth_adm2<-st_read("~/OneDrive - University of Texas at San Antonio//students//fikre/spatial_epi/ETH_adm2.shp")
```

```{r}
#merge dots to administrative data
eth_dots2000<-st_intersection(eth_dots, eth_adm2)

mapview(eth_dots["DHSCLUST"])+mapview(eth_adm2["NAME_2"])
```

## Read in DHS survey and recode stunting outcome

Next, I use the DHS survey data to estimate the prevalence of stunting in the ADM 2 regions.


```{r}
library(haven)
dhs2000<-read_dta("~/OneDrive - University of Texas at San Antonio//students//fikre/ethiopia_dhs/ETKR41DT/ETKR41FL.DTA")
dhs2000<-zap_labels(dhs2000)

library(car)
dhs2000$stunting<-ifelse(dhs2000$hw5/100<=-2&dhs2000$hw5/100!=-2,1,0)
#dhs2000$sex<-dhs2000$hc27

library(dplyr)
dhs2000<-dhs2000%>%
  mutate(wt = v005/1000000)%>%
  filter(complete.cases(stunting))%>%
  select(v001,stunting, wt, v021, v022)
```

## Merge survey data to sample locations

```{r}
dhs2000m<-merge(dhs2000, eth_dots2000, by.x="v001", by.y="DHSCLUST")

```


## Create survey estimates for new regions after spatial intersection

```{r}
library(survey)
options(survey.lonely.psu = "adjust")
des<-svydesign(ids = ~v021, strata = ~v022, weights = ~wt, data=dhs2000m)
names(dhs2000m)
est.stunt <- svyby(~stunting, ~ID_2, des, FUN=svymean, na.rm=T)

head(est.stunt)

```

## merge estimates to map and map stunting prevalence

```{r}
library(tigris)
library(mapview)
mdat<- geo_join(ethpoly, est.stunt, "ID_2","ID_2")
mapview(mdat["stunting"])


```