---
title: "DEM 5093/7093 Lab 2 using R - Map projections and transformations"
author: "Corey Sparks, Ph.D."
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
   html_document:
   toc: yes
   toc_float: yes
always_allow_html: yes
---

This lab complements the [exercise using Qgis for today ](https://docs.google.com/document/d/e/2PACX-1vQVSD4MKW4rTSQaNlM1cOULh-0-rSpoAaU3N_Xe7elujIj-ELgWTIGVDygF8qKl3QVBtpxPSi7YuDOS/pub)

Here, we use tidycensus to read some tract data, learn its projection information, transform it to a new coordinate system and measure some distance between features. 


```{r}
library(tidycensus)
library(sf)
library(dplyr)
```

### Read in Bexar county tracts

```{r, results="hide"}
sa_acs<-get_acs(geography = "tract",
                state="TX",
                county = c("Bexar"),
                year = 2017,
                variables=c( "DP05_0001E", 
                            "DP03_0119PE") ,
                geometry = T, output = "wide")

#create a county FIPS code - 5 digit
sa_acs$county<-substr(sa_acs$GEOID, 1, 5)

#rename variables and filter missing cases
sa_acs2<-sa_acs%>%
  mutate(totpop= DP05_0001E, ppov=DP03_0119PE) %>%
#  st_transform(crs = 102740)%>%
  na.omit()


# mydat<-sf::st_read("~/OneDrive - University of Texas at San Antonio/classes/gis_classwork/bexarpolygon.shp")
# plot(mydat)

```

### find coordinate system of current map
```{r}

st_crs(sa_acs2)
```



### create basic map
```{r}
library(tmap)
library(tmaptools)

tm_shape(sa_acs2)+
  tm_polygons("ppov", title="% in Poverty",
              palette="Blues",
              style="quantile",
              n=5 )+
  tm_format("World", 
            title="San Antonio Poverty Estimates\n- Quintile Breaks", legend.outside=T)+
  tm_scale_bar(position = c("left","bottom"))+
  tm_compass()


```


### re-project map into South Central Texas projection

```{r}
new_sa<-st_transform(sa_acs2, crs = 2278)

#Extract two tracts
twtr<-new_sa%>%
  filter(GEOID %in% c(48029181820, 48029110600))

# get centroid coordinates for two tracts
tr_co<-st_centroid(twtr)

#Measure feet apart
st_distance(tr_co)
```


# Using QGIS within R
This is another way to do the above task, by running a QGIS algorithm within R using the `qgisprocess` package. 

See the vignette [here](https://github.com/paleolimbot/qgisprocess) for more on what this package is and some examples.

```{r}
library(qgisprocess) #load the package

qgis_configure() #set up qgis - find the executable

# qgis_algorithms() lists all the available routines in QGIS
head(qgis_algorithms())

```

We can use `grep` to search for specific terms in the algorithms. 

```{r}
algs<-qgis_algorithms()

algs[grep(x = algs$algorithm, "distance"),"algorithm"]

qgis_show_help("qgis:distancematrix")
```

```{r}
out = qgis_run_algorithm(alg = "qgis:distancematrix",
               INPUT = tr_co[1,],
               INPUT_FIELD = "GEOID", 
               TARGET = tr_co[2,],
               TARGET_FIELD = "GEOID",
               MATRIX_TYPE = 0, 
               NEAREST_POINTS = 1)

output_sf <- sf::read_sf(qgis_output(out, "OUTPUT"))
output_sf$Distance

```