---
title: "Using IPUMS for Estimation of Population Characteristics in Various Geographic Areas"
author: "Corey Sparks, PhD"
date: "4/2/2020"
output:
   html_document:
    df_print: paged
    fig_height: 7
    fig_width: 7
    toc: yes
    includes:
      in_header: logo.html
---

In this example we will use the [IPUMS USA](https://usa.ipums.org/usa/) data to produce survey-based estimates for various geographic levels present in the IPUMS. This example uses the 2014-2018 ACS 5-year microdata. 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ipumsr)
ddi <- read_ipums_ddi("~/OneDrive - University of Texas at San Antonio/classes//gis_classwork/usa_00083.xml")
data <- read_ipums_micro(ddi)
data<-haven::zap_labels(data)

```

```{r, message=FALSE}
library(survey)
library(dplyr)
library(car)
library(ggplot2)
library(tigris)
library(classInt)
library(mapview)
```
### Download geographic data for Public Use Microdata Areas
```{r, results='hide'}
options(tigris_class = "sf")
pumas<-pumas(state = "TX", year = 2018, cb = T)
plot(pumas["GEOID10"], main = "Public Use Microdata Areas in Texas")
mapview(pumas)
```

```{r}
names(data)<-tolower(names(data))
```

## Prepare variables
Here I recode several demographic variables
```{r}
#weight variables
data$pwt <- data$perwt/100
data$hwt <- data$hhwt/100

data$hisp <- Recode(data$hispan, recodes = "9=NA; 1:4='Hispanic'; 0='NonHispanic'")
data$race_rec <- Recode(data$race, recodes = "1='White'; 2='Black'; 3='Other'; 4:6='Asian'; 7:9='Other'")
data$race_eth <- interaction(data$hisp, data$race_rec, sep = "_")
data$race_eth  <- as.factor(ifelse(substr(as.character(data$race_eth),1,8) == "Hispanic", "Hispanic", as.character(data$race_eth)))
data$race_eth <- relevel(data$race_eth, ref = "NonHispanic_White")

data$male <- ifelse(data$sex == 1,1,0)

data$educ_level<- Recode(data$educd, recodes = "2:61='0LT_HS';62:64='1_HSD/GED';65:80='2_somecoll';90:100='2_somecoll'; 81:83='3_AssocDegree';101='4_bachelordegree'; 110:116='4_BAplus_GradDegree'; else=NA")

data$employed <- Recode(data$wrklstwk, recodes = "1=0;2=1; else=NA")

data$cit<-Recode(data$citizen, recodes = "1='US born'; 2='naturalized'; 3:4='notcitizen';else=NA ")

data$ind_group<-Recode(data$ind, recodes = "170:490='ag_extract'; 770='construction'; 1070:3990='manufac'; 4070:5790='whole_retail'; 6070:6390='trans'; 6470:6780='information'; 6870:7190= 'fire'; 7270=7790='prof/sci/manage'; 7860:8470='edu/social'; 8560:8690='arts'; 8770:9290='other'; 9370:9590='public_adm'; 9670:9870='military'; else=NA ")

data$proftech <- Recode(data$ind, recodes = "7270:7490=1; 0=NA; else=0")

data$agecat<-cut(data$age, breaks = c(0, 18, 20, 30, 40, 50, 65, 120), include.lowest = T)

#data$edu_scal_inc <- ave(data$incwage, data$male, FUN = scale)
```

## Generate survey design object
Here we identify the person weights and the survey design variables.

```{r}
des<-svydesign(ids=~cluster, strata=~ strata, weights = ~pwt, data=data)
```

## perform survey estimation for PUMAs
The `svyby()` function allows us calculate estimates for different **sub-domains** within the data, this could be a demographic characteristic, but we'll use our geographic level. 

```{r}
puma_est_edu<-svyby(formula = ~educ_level, by = ~puma,design=subset(des,age>25), FUN=svymean, na.rm=T )
puma_est_employ<-svyby(formula = ~employed, by = ~puma,design=subset(des, age%in%18:65), FUN=svymean, na.rm=T )
puma_est_industry<-svyby(formula = ~proftech, by = ~puma,design=subset(des, employed==1), FUN=svymean, na.rm=T )

head(puma_est_edu)
head(puma_est_employ)
head(puma_est_industry)
```

## join to geography
```{r}
pumas$puma<-as.numeric(pumas$PUMACE10)
geo1<-geo_join(pumas, puma_est_employ, "puma", "puma")
head(geo1)

geo2<-geo_join(pumas, puma_est_industry, "puma", "puma")
head(geo2)

```

## Map estimates
```{r}
geo1%>%
  mutate(emp_cut=cut(100*employed,breaks = data.frame(classIntervals(var=100*puma_est_employ$employed, n=5, style="jenks")[2])[,1],include.lowest = T))%>%
  ggplot()+geom_sf(aes(fill=emp_cut, color=emp_cut))+
  scale_fill_brewer(palette = "Blues") + 
  scale_color_brewer(palette = "Blues")+
  ggtitle(label="Employment rate in Texas PUMAs", subtitle = "ACS 2014 - 2018")

geo1%>%
  mutate(emp_cut=cut(100*employed,breaks = data.frame(classIntervals(var=100*puma_est_employ$employed, n=5, style="jenks")[2])[,1],include.lowest = T))%>%
  filter(grepl(geo1$NAME10, pattern = "San Antonio")==T)%>%
  ggplot()+geom_sf(aes(fill=emp_cut), color="black")+
  scale_fill_brewer(palette = "Blues") + 
  scale_color_brewer(palette = "Blues")+
  ggtitle(label="Employment rate in San Antonio PUMAs", subtitle = "ACS 2014 - 2018")

mapview(geo1, zcol="employed")
```

```{r}
geo2%>%
  mutate(p_cut=cut(100*proftech,breaks = data.frame(classIntervals(var=100*puma_est_industry$proftech, n=5, style="jenks")[2])[,1],include.lowest = T))%>%
  ggplot()+geom_sf(aes(fill=p_cut))+
  scale_fill_brewer(palette = "Blues") + 
  scale_color_brewer(palette = "Blues")+
  ggtitle(label="Percent in Professional/Technical Jobs  \n in Texas PUMAs", subtitle = "ACS 2014 - 2018")

geo2%>%
mutate(p_cut=cut(100*proftech,breaks = data.frame(classIntervals(var=100*puma_est_industry$proftech, n=5, style="jenks")[2])[,1],include.lowest = T))%>%
  filter(grepl(geo1$NAME10, pattern = "San Antonio")==T)%>%
  ggplot()+geom_sf(aes(fill=p_cut), color="black")+
  scale_fill_brewer(palette = "Blues") + 
  scale_color_brewer(palette = "Blues")+
  ggtitle(label="Percent in Professional/Technical Jobs  \n in San Antonio PUMAs", subtitle = "ACS 2014 - 2018")


```


## Estimation for metro areas
Here we use core based statistical areas instead of PUMAs

```{r, results='hide'}
mets<-core_based_statistical_areas(cb = T, year = 2018)
mets<-mets[grep(mets$NAME,pattern =  "TX"),]
plot(mets["NAME"])

sts<-states(cb=T, year=2018)
sts<-sts%>%
  filter(GEOID==48)
```
## estimates by metro area
```{r}
met_est_edu<-svyby(formula = ~educ_level, by = ~met2013,design=subset(des,age>25), FUN=svymean, na.rm=T )
met_est_employ<-svyby(formula = ~employed, by = ~met2013,design=subset(des, age%in%18:65), FUN=svymean, na.rm=T )
met_est_industry<-svyby(formula = ~proftech, by = ~met2013,design=subset(des, employed==1), FUN=svymean, na.rm=T )

head(met_est_edu)
head(met_est_employ)
head(met_est_industry)

```


```{r}
mets$met2013<-as.numeric(mets$GEOID)
geo3<-geo_join(mets, met_est_employ,by_sp= "met2013",by_df= "met2013")

geo3%>%
  mutate(emp_cut=cut(100*employed,breaks = data.frame(classIntervals(var=100*met_est_employ$employed, n=5, style="jenks")[2])[,1],include.lowest = T))%>%
  ggplot()+
  geom_sf(aes(fill=emp_cut))+
  scale_fill_brewer(palette = "Blues", na.value = "grey50") + 
  scale_color_brewer(palette = "Blues",na.value = "grey50")+
  ggtitle(label="Employment rate in Texas \nMetro Areas", subtitle = "ACS 2014 - 2018")
  

mapview(geo3, zcol="employed")
```


## Estimation for Counties
## 
```{r, results='hide'}
cos<-counties(cb= T,state = "TX", year = 2018)
plot(cos["NAME"])

sts<-states(cb=T, year=2018)
sts<-sts%>%
  filter(GEOID==48)
```
## estimates by county area
```{r}
cos_est_edu<-svyby(formula = ~educ_level, by = ~countyfip,design=subset(des,age>25), FUN=svymean, na.rm=T )
cos_est_employ<-svyby(formula = ~employed, by = ~countyfip,design=subset(des, age%in%18:65), FUN=svymean, na.rm=T )
cos_est_industry<-svyby(formula = ~proftech, by = ~countyfip,design=subset(des, employed==1), FUN=svymean, na.rm=T )

head(cos_est_edu)
head(cos_est_employ)
head(cos_est_industry)

```


```{r}
cos$cofip<-as.numeric(cos$COUNTYFP)
geo4<-geo_join(cos, cos_est_employ,by_sp= "cofip",by_df= "countyfip")

geo4%>%
  mutate(emp_cut=cut(100*employed,breaks = data.frame(classIntervals(var=100*cos_est_employ$employed, n=5, style="jenks")[2])[,1],include.lowest = T))%>%
  ggplot()+
  geom_sf(aes(fill=emp_cut))+
  scale_fill_brewer(palette = "Blues", na.value = "grey50") + 
  scale_color_brewer(palette = "Blues",na.value = "grey50")+
  ggtitle(label="Employment rate in Texas \nCounties", subtitle = "ACS 2014 - 2018")
  


```
