---
title: "DEM 5093/7093 Lab 1 using  R - Basic Map making"
author: "Corey Sparks, Ph.D."
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
always_allow_html: yes
---

### Get a Census developer API Key
Obtain one at http://api.census.gov/data/key_signup.html

### Save your API key to your working directory
use `tidycensus::census_api_key(key =  "yourkeyhere", install = T)`

one time to install your key for use in `tidycensus`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, eval=TRUE}
library(tidycensus)
library(tidyverse)
library(sf)
library(ggplot2)
```

```{r}
v15_Profile <- load_variables(2017 , "acs5/profile", cache = TRUE) #demographic profile tables
View(v15_Profile)


#Search for variables by keywords in the label 
v15_Profile[grep(x = v15_Profile$label, "POVERTY"), c("name", "label")]
v15_Profile[grep(x = v15_Profile$label, "Built 2000 to 2009"), c("name", "label")]
```

\newpage
## Extract from ACS summary file data

Here I get data profile variables from 2017 for Bexar County, TX Census Tracts

The data profile tables are very useful because they contain lots of pre-calculated variables.

Here is a query where we extract several variables from tehe 2017 ACS for Bexar County, Texas. We can also get the spatial data by requesting `geometry=TRUE`.  Useing `output="wide"` will put each variable in a column of the data set, with each row being a census tract. 

```{r, results='hide'}
sa_acs<-get_acs(geography = "tract",
                state="TX",
                county = c("Bexar"),
                year = 2017,
                variables=c( "DP05_0001E", "DP03_0119PE") ,
                geometry = T, output = "wide")

#create a county FIPS code - 5 digit
sa_acs$county<-substr(sa_acs$GEOID, 1, 5)

#rename variables and filter missing cases
sa_acs2<-sa_acs%>%
  mutate(totpop= DP05_0001E, ppov=DP03_0119PE) %>%
#  st_transform(crs = 102740)%>%
  na.omit()

```

## Write data out to shapefile
```{r, message=FALSE, warning=FALSE, eval=F}
#change the directory
sf::st_write(sa_acs2,dsn="C:/Users/ozd504/OneDrive - University of Texas at San Antonio/classes/gis_classwork",layer="sa_tr_dp03", driver="ESRI Shapefile")

```

\newpage

## Some basic mapping of variables
Here I generate a quantile break for % black in census tracts and compare it to a Jenks break
```{r, message=FALSE, fig.width=12, fig.height=8}

library(classInt)
library(patchwork)
library(dplyr)

ppov_map<-sa_acs2 %>%
  mutate(cpov=cut(ppov,breaks = quantile(ppov, na.rm=T, p=seq(0,1,length.out = 6)),include.lowest = T),
           jpov = cut(ppov,breaks=data.frame(classIntervals(var=sa_acs2$ppov, n=5, style="pretty")[2])[,1], include.lowest = T))

library(ggsn)

p1<-ggplot(ppov_map, aes(fill = cpov)) + 
  geom_sf(color="grey50") + 
  ggtitle("Proportion in poverty", 
          subtitle = "Bexar County Texas, 2017 - Quantile Breaks")+
    scale_fill_brewer(palette = "Reds") + 
  scale_color_brewer(palette = "Reds")+
  theme(axis.text.x = element_blank(), axis.text.y = element_blank())+
  north(ppov_map)+
  scalebar(ppov_map, location="bottomleft", dist=5, transform = T,dist_unit = "km",  model="WGS84", st.size =2 )
p1



p2<-ggplot(ppov_map, aes(fill = jpov)) + 
  geom_sf(color="grey50") + 
  ggtitle("Proportion in Poverty", 
          subtitle = "Bexar County Texas, 2017 - Jenks Breaks")+
  scale_fill_brewer(palette = "Reds") + 
  scale_color_brewer(palette = "Reds")+
    theme(axis.text.x = element_blank(), axis.text.y = element_blank())+
  north(ppov_map)+
  scalebar(ppov_map, location="bottomleft", dist=5, transform = T,dist_unit = "km",  model="WGS84", st.size =2)
p2
  
p1+ p2

```

```{r, eval=FALSE}
ggsave(filename="C:/Users/ozd504/OneDrive - University of Texas at San Antonio//classes/gis_classwork/lab1map1.png")

```

### Slightly easier way using `tmap`

```{r}
library(tmap)
library(tmaptools)

tm_shape(sa_acs2)+
  tm_polygons("ppov", title="% in Poverty", palette="Blues", style="quantile", n=5 )+
  tm_format("World", title="San Antonio Poverty Estimates - Quantile Breaks", legend.outside=T)+
  tm_scale_bar()+
  tm_compass()

tm_shape(sa_acs2)+
  tm_polygons("ppov", title="% in Poverty", palette="Blues", style="jenks", n=5,legend.hist=T )+
  tm_format("World", title="San Antonio Poverty Estimates - Jenks Breaks", legend.outside=T)+
  tm_scale_bar()+
  tm_compass()

```


## Interactive map with mapview

```{r}
library(mapview)
library(RColorBrewer)
ppov_map$jpov<-relevel(ppov_map$jpov,ref = "[0,10]" )
pal <- colorRampPalette(brewer.pal(7, "Blues")) #set colors
mapview(ppov_map, zcol="jpov",  legend=T,map.types="OpenStreetMap", layer.name="% in Poverty")


```



