---
title: "DEM 5093/7093 Lab 1 using  R - Basic Map making"
author: "Corey Sparks, Ph.D."
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
always_allow_html: yes
---

### Get a Census developer API Key
Obtain one at the Census Developer [website](http://api.census.gov/data/key_signup.html)

### Save your API key to your working directory
use `tidycensus::census_api_key(key =  "yourkeyhere", install = T)`

One time to install your key for use in `tidycensus`

```{r, message=FALSE, eval=TRUE}
library(tidycensus)
library(tidyverse)
library(sf)
library(ggplot2)
library(classInt)
```


### Examine data profile tables
The `load_variables()` function will load all available variables in the ACS for a given year
```{r}
v15_Profile <- load_variables(year = 2019 ,
                              dataset =  "acs5/profile",
                              cache = TRUE)


```

Calling `View(v15_Profile)` will let you interactively browse and filter the ACS variables, this is one way to search for what you're looking for.


### Search for variables by keywords in the label 
```{r}
v15_Profile%>%
  filter(grepl(pattern = "POVERTY", x = label))%>%
  select(name, label)
```

```{r}
v15_Profile%>%
  filter(grepl(pattern = "Built 2000 to 2009", x = label))%>%
  select(name, label)

```

\newpage
## Extract from ACS summary file data

The `tidycensus` package has a function `get_acs()` that will download data from the Census API for you automatically **assuming you've installed your key from above** 

Here I get data profile variables from 2017 for Bexar County, TX Census Tracts


Here is a query where we extract several variables from the 2017 ACS for Bexar County, Texas. We can also get the spatial data by requesting `geometry=TRUE`. 

Using `output="wide"` will put each variable in a column of the data set, with each row being a census tract. 

```{r, results='hide'}
sa_acs<-get_acs(geography = "tract",
                state="TX",
                county = "Bexar",
                year = 2019,
                variables=c("DP05_0001E", "DP03_0119PE") ,
                geometry = T,
                output = "wide")

head(sa_acs)
```

Here, I create some other variables that we may need later
```{r}
# create a county FIPS code - 5 digit
sa_acs$county<-substr(sa_acs$GEOID, 1, 5)

# rename variables and filter missing cases
sa_acs2<-sa_acs%>%
  mutate(totpop= DP05_0001E,
         ppov=DP03_0119PE) %>%
  st_transform(crs = 2919)%>%
  na.omit()

```

## Write data out to shapefile
You may need to create or modify some data in R and then use it in the desktop GIS (QGIS), we can write any data from R into a variety of data formats using the  `sf::st_write()` function.


```{r, message=FALSE, warning=FALSE, eval=FALSE}
#change the directory for your computer
sf::st_write(sa_acs2,
             dsn="../data/sa_tr_dp03.gpkg",
             layer="sa_tr_dp03",
             driver="GPKG") ## Save as geopackage format - QGIS likes this

```

\newpage

## Some basic mapping of variables
Here I generate a quantile break for % black in census tracts and compare it to a Jenks break. *Note* in ggplot, the Jenks break is harder to do
```{r, message=FALSE, fig.width=12, fig.height=8}

library(patchwork)
library(ggsn)
library(XploreR)
source("https://raw.githubusercontent.com/coreysparks/Rcode/master/mutate_map_funs.R")

ppov_map<-sa_acs2 %>%
  mutate_map_brks(ppov, n=6, style="quantile")%>%
           mutate_map_brks(ppov, n=6, style="jenks")



p1<-ggplot(ppov_map, aes(fill = ppov_qbrks)) + 
  geom_sf(color="grey50") + 
  ggtitle("Proportion in poverty", 
          subtitle = "Bexar County Texas, 2017 - Quantile Breaks")+
    scale_fill_brewer(palette = "Reds") + 
  scale_color_brewer(palette = "Reds")+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())+
  north(ppov_map)+
  scalebar(ppov_map, location="bottomleft",
           dist=5, transform = T,
           dist_unit = "km",
           model="WGS84",
           st.size =2 )
p1



p2<-ggplot(ppov_map, aes(fill = ppov_jbrks)) + 
  geom_sf(color="grey50") + 
  ggtitle("Proportion in Poverty", 
          subtitle = "Bexar County Texas, 2017 - Jenks Breaks")+
  scale_fill_brewer(palette = "Reds") + 
  scale_color_brewer(palette = "Reds")+
    theme(axis.text.x = element_blank(),
          axis.text.y = element_blank())+
  north(ppov_map)+
  scalebar(ppov_map, 
           location="bottomleft",
           dist=5,
           transform = T,
           dist_unit = "km",
           model="WGS84",
           st.size =2)
p2
  
p1+ p2

```

You can save the image from above to your computer by using `ggsave()`
```{r, eval=FALSE}
ggsave(filename="../images/lab1map1.png",
       dpi = "print")

```

### Slightly easier way using `tmap`

The `tmap` package is an attractive alternative to using `ggplot()` when making maps, and makes basic cartographic principles easier.

*Note* `style="fisher"` is equivalent to `style="jenks"` and scales better to larger data.

```{r}
library(tmap)
library(tmaptools)

tm_shape(sa_acs2)+
  tm_polygons("ppov",
              title="% in Poverty",
              palette="Blues",
              style="quantile", n=5 )+
  tm_format("World",
            title="San Antonio Poverty Estimates - Quantile Breaks",
            legend.outside=T)+
  tm_scale_bar()+
  tm_compass()

tm_shape(sa_acs2)+
  tm_polygons("ppov",
              title="% in Poverty",
              palette="Blues",
              style="fisher",
              n=5,
              legend.hist=T )+
  tm_format("World",
            title="San Antonio Poverty Estimates - Jenks Breaks",
            legend.outside=T)+
  tm_scale_bar()+
  tm_compass()

```


## Interactive map with mapview

```{r}
library(mapview)
library(RColorBrewer)
ppov_map$ppov_jbrks<-relevel(ppov_map$ppov_jbrks,ref = "0-5.5" )

pal <- colorRampPalette(brewer.pal(7, "Blues")) #set colors

mapview(ppov_map,
        zcol="ppov_jbrks",
        legend=T,
        map.types="OpenStreetMap",
        layer.name="% in Poverty")


```



