---
title: "DEM 5093/7093 Point Pattern Analysis"
author: "Corey Sparks, Ph.D."
date:  "`r format(Sys.time(), '%d %B, %Y')`"
output:
   html_document:
    df_print: paged
    fig_height: 7
    fig_width: 7
    toc: yes
    toc_float: yes
---

This example shows how to use R and QGIS from within R to perform a series of common point pattern analysis techniques. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(mapview)
library(sf)
library(censusxy)
library(dplyr)
```

```{r}
addr<-read.csv(url("https://raw.githubusercontent.com/coreysparks/data/master/wic_west_side.csv"))
addr<-addr[c(6, 12:14)]
names(addr)<-c("street", "city", "st", "zip")
head(addr)

results<-cxy_geocode(addr,
                     street = "street",
                     city = "city",
                     state ="st",
                     zip = "zip",
                     class="sf",
                     output = "simple")
results.proj<-st_transform(results,
                           crs = 2278)
```


```{r}
mapview(results)
```


# mean feature - average of coordinates

```{r}
mean_feature<-apply(st_coordinates(results.proj), MARGIN = 2, FUN = mean)
mean_feature<-data.frame(place="meanfeature", x=mean_feature[1], y= mean_feature[2])
mean_feature<-st_as_sf(mean_feature, coords = c("x", "y"), crs= 2278)

mapview(mean_feature, col.regions="red")+mapview( results)
```

## Central feature - Median of coordinates
```{r}
median_feature<-apply(st_coordinates(results.proj), MARGIN = 2, FUN = median)
median_feature<-data.frame(place="medianfeature", x=median_feature[1], y= median_feature[2])
median_feature<-st_as_sf(median_feature, coords = c("x", "y"), crs= 2278)

mapview(median_feature, col.regions="green")+
  mapview(mean_feature, col.regions="red")+
  mapview( results)

```


## Buffer points
```{r}
wicbuff<- st_buffer(results.proj, dist = 2500)
mapview(wicbuff)+mapview(results.proj, col.regions="green")

```

## Convex hull plot
```{r}
chull <- st_convex_hull(st_union(results))

mapview(chull)+
  mapview(results, col.regions = "green")
```


## kernel density  - You need projected data for this to work right
R can do kernel density maps, but using simple features it's kind of complicated. I will use Qgis through R instead using the `qgisprocess` [package](https://github.com/paleolimbot/qgisprocess)

```{r}
library(qgisprocess)
qgis_configure()

```
To use this, we need to find the name of the Qgis algorithm we want. `qgis_algorithms()` can return all available algorithms, then we can either filter it with `View()` or use grep to search for one. 

```{r}

algs<-qgis_algorithms()
algs[grepl(pattern = "density", x = algs$algorithm ),]
qgis_show_help("qgis:heatmapkerneldensityestimation")


```

Run the algorithm
```{r, message=FALSE, warning=FALSE, results='hide'}
wic_dens<-qgis_run_algorithm(algorithm ="qgis:heatmapkerneldensityestimation",
         INPUT=results.proj,
         RADIUS = 5000,
         PIXEL_SIZE = 100,
         KERNEL = 0,
         OUTPUT=file.path(getwd(), "wicdenst.TIF"),
         load_output = TRUE)
library(raster)
library(RColorBrewer)

result<- qgis_as_raster(wic_dens)

projection(result)<-crs(results.proj)
mapview(result)+mapview(results.proj)
```

## Spatial join
A spatial join can combine attributes of one layer with another layer. Here I combine census variables with the WIC clinic points.

```{r, results="hide"}
library(tidycensus)
library(dplyr)
#load census tract data
sa_acs<-get_acs(geography = "tract",
                state="TX",
                county = "Bexar", 
                year = 2019,
                variables=c( "DP05_0001E", "DP03_0009P", "DP03_0062E", "DP03_0119PE",
                           "DP05_0001E","DP02_0009PE","DP02_0008PE","DP02_0040E","DP02_0038E",
                            "DP02_0066PE","DP02_0067PE","DP02_0080PE","DP02_0092PE",
                        "DP03_0005PE","DP03_0028PE","DP03_0062E","DP03_0099PE","DP03_0101PE",
                            "DP03_0119PE","DP04_0046PE","DP05_0072PE","DP05_0073PE",
                            "DP05_0066PE", "DP05_0072PE", "DP02_0113PE") ,
                geometry = T, output = "wide")

#rename variables and filter missing cases
sa_acs2<-sa_acs%>%
  mutate(totpop= DP05_0001E, pwhite=DP05_0072PE, 
         pblack=DP05_0073PE , phisp=DP05_0066PE,
         phsormore=DP02_0066PE,punemp=DP03_0009PE, medhhinc=DP03_0062E,
         ppov=DP03_0119PE)%>%
  dplyr::select(GEOID, totpop, pblack, pwhite, phisp, punemp, medhhinc, ppov)

sa_acs2<-st_transform(sa_acs2, crs = 2278)
sa_trol<-st_cast(sa_acs2, "MULTILINESTRING")
```

```{r}
spjoin<-st_join(results.proj, sa_acs2)
head(spjoin)

mapview(spjoin["punemp"])+mapview(sa_trol)
```

## Count points in polygons
Point in polygon operations are actually a spatial intersection (more on this next week!) where we see how many points fall within a given polygon. 

```{r}
sa_acs2$nwic<- lengths(st_intersects(sa_acs2, results.proj))
mapview(sa_acs2, zcol="nwic")+
  mapview(results.proj, col.regions = "green")
```


## Thiessen/Voronoi Polygons
Thiessen or Voronoi polygons are a process where we can convert points into polygons.

```{r}
algs[grepl(pattern = "voronoi", x = algs$algorithm ),]
qgis_show_help("qgis:voronoipolygons")
```
```{r, message=FALSE, warning=FALSE, results='hide'}


wic_von<-qgis_run_algorithm(alg="qgis:voronoipolygons",
         INPUT=results.proj,
         OUTPUT=file.path(tempdir(), "wicvon.shp"),
         load_output = TRUE)

wic_von<-sf::read_sf(qgis_output(wic_von, "OUTPUT"))

mapview(wic_von, alpha=.75)+
  mapview(results.proj, col.regions="green")

```


## Nearest Neighbor analysis

```{r, message=FALSE, warning=FALSE, results='hide'}
library(spatstat)
wic.pp<-as.ppp(as(results.proj, "Spatial"))

plot(nearest.neighbour(wic.pp))
```

```{r}
algs[grepl(pattern = "nearest", x = algs$algorithm ),]
qgis_show_help("native:nearestneighbouranalysis")
```



```{r}


wic_nn<-qgis_run_algorithm(alg="native:nearestneighbouranalysis",
         INPUT=results.proj,
        OUTPUT_HTML_FILE=file.path(tempdir(), "wicnn.html"),
         load_output = TRUE)


```