---
title: "Intro to Point Pattern Analysis using R and QGIS"
author: "Corey Sparks, Ph.D."
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
   html_document:
    df_print: paged
    fig_height: 7
    fig_width: 7
    toc: yes
    toc_float: yes
always_allow_html: yes
download_code: yes    
---

This example shows how to use R and QGIS from within R to perform a series of common point pattern analysis techniques. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(mapview)
library(sf)
library(censusxy)
library(dplyr)
```
## Create point layer from grocery data from AtoZdatabase
#### just use the lat / long information in the data!
```{r}
addr<-read.csv(url("https://raw.githubusercontent.com/coreysparks/DEM7093/main/data/west_side_groceries.csv"))
addr <- addr%>%
  dplyr::select(names(addr)[c(4,6,14,23:24,41,42)])

results <- st_as_sf(addr, coords=c("Longitude", "Latitude"), crs=4269,agr="constant")

results.proj<-st_transform(results,
                           crs = 2278)

```




```{r}
mv1 <- mapview(results.proj)
mapshot(mv1, file = paste0(getwd(), "/map1.png"))


mv1
```

```{r}
library(tmap)
library(tmaptools)
library(OpenStreetMap)
bg<-  read_osm(results.proj, ext=1.1)

tmap_mode("view")
tm_basemap("OpenStreetMap.Mapnik")+
  tm_shape(results.proj)+
  tm_dots()

tmap_mode("plot")
```

```{r}
library(ggplot2)
library(ggmap)
library(ggspatial)
library(leaflet)
leaflet() %>%
  addTiles()%>%
  addMarkers(lng = addr$Longitude, lat= addr$Latitude)

```



# mean feature - average of coordinates

```{r}
mean_feature<-apply(st_coordinates(results.proj), MARGIN = 2, FUN = mean)
mean_feature<-data.frame(place="meanfeature", x=mean_feature[1], y= mean_feature[2])
mean_feature<-st_as_sf(mean_feature, coords = c("x", "y"), crs= 2278)


tmap_mode("view")
tm_basemap("OpenStreetMap.Mapnik")+
  tm_shape(results.proj, size = .2)+
  tm_dots(col = "green")+
  tm_shape(mean_feature)+
  tm_dots(col = "red", size = .2)


```




## Central feature - Median of coordinates
```{r}
median_feature<-apply(st_coordinates(results.proj),
                      MARGIN = 2,
                      FUN = median)

median_feature<-data.frame(place="medianfeature",
                           x=median_feature[1],
                           y= median_feature[2])

median_feature<-st_as_sf(median_feature,
                         coords = c("x", "y"),
                         crs= 2278)


tmap_mode("view")
tm_basemap("OpenStreetMap.Mapnik")+
  tm_shape(results.proj)+
  tm_dots(col = "green", size = .2)+
  tm_shape(mean_feature)+
  tm_dots(col = "red", size=.2)+
  tm_shape(median_feature)+
  tm_dots(col = "blue", size = .2)

```



## Buffer points
```{r}
grocery_buff<- st_buffer(results.proj, dist = 1000)

#tmap_mode("plot")
tm_basemap("OpenStreetMap.Mapnik" )+
  tm_shape(results.proj, is.master = T)+
  tm_dots(col = "green")+
  tm_shape(grocery_buff)+
  tm_polygons(col = "red", alpha = .1)


```


## Convex hull plot
```{r}
chull <- st_convex_hull(st_union(results))

tmap_mode("view")
tm_basemap("OpenStreetMap.Mapnik")+
  tm_shape(results.proj)+
  tm_dots(col = "green")+
  tm_shape(chull)+
  tm_polygons(col = "grey", alpha = .5)

```


## kernel density  - 
### You need projected data for this to work right

R can do kernel density maps, but using simple features it's kind of complicated. Here is one way to do this using the `SpatialKDE` package:

```{r, message=FALSE, warning=FALSE}

library(SpatialKDE)

grid_groc <- results.proj %>%
  create_grid_rectangular(cell_size = 1000, side_offset = 2000)

kde <- results.proj%>%
  kde(band_width = 2000, kernel= "quartic", grid = grid_groc)

tm_shape(kde)+
  tm_polygons(col = "kde_value", palette= "viridis", title = "Density Estimate")+
  tm_shape(results.proj)+
  tm_dots()

```

Alternatively, you can use Qgis through R instead using the `qgisprocess` [package](https://github.com/paleolimbot/qgisprocess)

```{r}
library(qgisprocess)
qgis_configure()

```
To use this, we need to find the name of the Qgis algorithm we want. `qgis_algorithms()` can return all available algorithms, then we can either filter it with `View()` or use grep to search for one. 

```{r}

algs<-qgis_algorithms()
algs[grepl(pattern = "density", x = algs$algorithm ),]
qgis_show_help("qgis:heatmapkerneldensityestimation")


```

Run the algorithm


```{r}

wic_dens<-qgis_run_algorithm(algorithm ="qgis:heatmapkerneldensityestimation",
         INPUT=results.proj,
         RADIUS = 2000,
         PIXEL_SIZE = 100,
         KERNEL = 0,
         OUTPUT=file.path(getwd(), "wicdenst.TIF"),
         load_output = TRUE)
library(raster)
library(RColorBrewer)

result<- qgis_as_raster(wic_dens)

projection(result)<-crs(results.proj)

tmap_mode("view")
tm_basemap("OpenStreetMap.Mapnik")+
  tm_shape(result)+
  tm_raster()+
  tm_shape(results.proj, is.master = T)+
  tm_dots(col="red")
  

```


## Spatial join
A spatial join can combine attributes of one layer with another layer. Here I combine census variables with the WIC clinic points.

```{r, results="hide"}
library(tidycensus)

#load census tract data
sa_acs<-get_acs(geography = "tract",
                state="TX",
                county = "Bexar", 
                year = 2019,
                variables=c( "DP05_0001E", "DP03_0062E",
                             "DP03_0119PE","DP02_0066PE",
                             "DP03_0062E","DP03_0119PE",
                             "DP05_0073PE","DP05_0066PE") ,
                geometry = T, output = "wide")

#rename variables and filter missing cases
sa_acs2<-sa_acs%>%
  mutate(totpop= DP05_0001E,
         pblack=DP05_0073PE,
         phisp=DP05_0066PE,
         phsormore=DP02_0066PE,
         medhhinc=DP03_0062E,
         ppov=DP03_0119PE)%>%
  dplyr::select(GEOID, totpop, pblack, phisp, medhhinc, ppov)

sa_acs2<-st_transform(sa_acs2, crs = 2278)
sa_trol<-st_cast(sa_acs2, "MULTILINESTRING") #make outline
```

```{r}
spjoin<-st_join(results.proj, sa_acs2)
#head(spjoin)

tmap_mode("view")
tm_basemap("OpenStreetMap.Mapnik")+
  tm_shape(spjoin, is.master = T)+
  tm_dots("ppov", size = .1)+
  tm_shape(sa_acs2)+
  tm_polygons(alpha = .1)


```


## Count points in polygons
Point in polygon operations are actually a spatial intersection (more on this next week!) where we see how many points fall within a given polygon. 

```{r}
sa_acs2$ngroc<- lengths(st_intersects(sa_acs2, results.proj))

sa_acs2$groc_pc <- 1000*(sa_acs2$ngroc/sa_acs2$totpop)

tmap_mode("plot")
m<-tm_basemap("OpenStreetMap.Mapnik")+
  tm_shape(sa_acs2)+
  tm_polygons("groc_pc")+
  tm_shape(spjoin, is.master = T)+
  tm_dots( size = .01)

m
```



## Thiessen/Voronoi Polygons
Thiessen or Voronoi polygons are a process where we can convert points into polygons. I use QGIS to do this, R has issues. 

```{r}
algs[grepl(pattern = "voronoi", x = algs$algorithm ),]
qgis_show_help("qgis:voronoipolygons")
```
```{r, message=FALSE, warning=FALSE, results='hide'}


wic_von<-qgis_run_algorithm(alg="qgis:voronoipolygons",
         INPUT=results.proj,
         OUTPUT=file.path(tempdir(), "wicvon.shp"),
         load_output = TRUE)

wic_von<-sf::read_sf(qgis_output(wic_von, "OUTPUT"))

tmap_mode("view")
map1<-tm_basemap("OpenStreetMap.Mapnik")+
  tm_shape(wic_von)+
  tm_polygons(col="grey", alpha=.4)+
  tm_shape(results.proj)+
  tm_dots( size = .01)
 

# plot map
map1
```


## Nearest Neighbor analysis
Nearest neighbor analysis is used to test if a pattern of points is distrubuted randomly or clustered.

```{r, message=FALSE, warning=FALSE, results='hide'}
library(spatstat)
wic.pp<-as.ppp(as(results.proj, "Spatial"))

plot(nearest.neighbour(wic.pp))
```

```{r}
algs[grepl(pattern = "nearest", x = algs$algorithm ),]
qgis_show_help("native:nearestneighbouranalysis")
```


Here I use QGIS to get a single test of the randomness

```{r}


wic_nn<-qgis_run_algorithm(alg="native:nearestneighbouranalysis",
         INPUT=results.proj,
        OUTPUT_HTML_FILE=file.path(tempdir(), "wicnn.html"),
         load_output = TRUE)

wic_nn
```
