---
title: "DEM 7093 - Some Raster analysis using R"
author: "Corey S. Sparks, PhD"
date: "June 20, 2018"
output: html_document
---

In this analysis, we use R to read in LandSat8 data, create a vegitation index and calculate zonal statistics of a raster. The data come from the [2006 National Land Cover Database](https://www.mrlc.gov/nlcd06_data.php).

First we relcassify the NLCD data into two classes, based on the value of the raster. In this case, I want to classify the pixels as to whether they are developed or undeveloped. 

We then use a census tract polygon layer to calculate the proportion of each tract's land area that is developed vs undeveloped. 


## Raster read in 
```{r }
library(rgdal)
library(raster)
library(gdalUtils)
nlcd<-raster("~/OneDrive - University of Texas at San Antonio/classes/dem7093/dem7093_21//data/nlcd16_48_lc/nlcd_2016_land_cover_48_20190424.img.vat.img")

 nlcdproj<-gdalwarp(srcfile ="~/OneDrive - University of Texas at San Antonio/classes/gis_classwork/nlcd16_48_lc/nlcd_2016_land_cover_change_index_48_20190424.img.vat.img", dstfile = "~/OneDrive - University of Texas at San Antonio/classes/gis_classwork/outputnlcd.tif",t_srs = nlcd , output_Raster = T)

 raster::plot(nlcdproj, main="Original Raster image")

```

## Clip raster to polygon
```{r, results='hide'}
library(sf)
library(tigris)
library(terra)
options(tigris_class = "sf")
counties<-tracts(state="TX", county = c("Bexar", "Comal"), cb=T, year=2016)
counties<-st_transform(counties, crs=2278)

# doesn't work

est<-terra::crop(nlcdproj, counties)
plot(est)

est2<-terra::mask(est, counties)
plot(est2)
# est<-mask(x=nlcdproj,mask=as_Spatial(counties))
# est2<-crop(x=nlcdproj, y=extent(as_Spatial(counties)))
# plot(est2)

```

## Raster reclassification:

```{r}
m<-c(1, 12, 0,13, 24, 1, 25, 95, 0)
rclmat<-matrix(m, ncol=3, byrow = T)
nlcdreclass<-raster::reclassify(x = est2, rcl = rclmat)

plot(nlcdreclass, main="Reclassified Raster image \n 1=developed, 0 = undeveloped",col= gray(100:0 / 100))

```


## Load the census tract polygons
```{r, results='hide'}
library(tigris)
satract<-tracts(state="48", county="029", year=2010)

library(sf)
satract<-st_as_sf(satract)
satract<-st_transform(satract, crs=  2278)

satract<-as(satract, "Spatial")
summary(satract)
```

## Calculate zonal statistics
Now we find the mean of the binary reclassified data, to estimate the percent of each tract's area that is developed. 

First we reproject the raster to the same coordinate system as the census data. I get the projection string from ^^^ above.

```{r}
nlcdreclass<-projectRaster(nlcdreclass, crs="+proj=lcc +lat_1=20 +lat_2=60 +lat_0=40 +lon_0=-96 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0")

satract$pct_developed<-as.numeric(extract(nlcdreclass, satract, fun=mean)) #this takes a minute or two

satract<-st_as_sf(satract)

library(ggplot2)
library(dplyr)
satract<-satract%>%
  filter(complete.cases(pct_developed))

tractlines<-st_boundary(satract)

```

```{r}
library(tmap)
tm_shape(satract)+
  tm_polygons("pct_developed",style="quantile", n=5,legend.hist=T )+
  tm_shape(tractlines)+
  tm_lines(col="black")+
  tm_format("World",
            title="San Antonio Developed Land Area",
            legend.outside=TRUE )

```
```{r}
library(classInt)
m1<-satract%>%
  mutate(devcut=cut(satract$pct_developed,breaks=data.frame(classIntervals(var=satract$pct_developed, n=5, style="jenks")[2])[,1], include.lowest = T))%>%
  ggplot(aes( fill=devcut))+geom_sf()+
  scale_fill_brewer(palette = "Blues") + 
  scale_color_brewer(palette = "Blues")+
  geom_sf(data=tractlines,fill=NA, color="black")+
  guides(fill=guide_legend(title="Percent Developed"))+ggtitle(label="Percent of Land Area Developed, 2006", subtitle ="NLCD 2006")+theme(axis.text.x = element_blank(), axis.text.y = element_blank())+theme( legend.text = element_text(size = 14), plot.title = element_text(size = 18), strip.text.x = element_text(size = 16))

m1
```

I used this method in a [paper](https://www.sciencedirect.com/science/article/pii/S1877584511000529) I wrote in 2013, although in that paper I did the diversity of the land cover. That could be done like:

```{r}
#nlcdproj<-projectRaster(nlcd, crs="+proj=lcc +lat_1=20 +lat_2=60 +lat_0=40 +lon_0=-96 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0")

library(vegan)

shannonVegan <- function(x, ...) {
    diversity(table(x), index="shannon")
}

#shanVegOut <- focal(nlcdproj, fun=shannon, pad=T)  

satract$var_land<-as.numeric(extract(nlcdproj, satract, fun=shannonVegan)) #this takes a minute or two

```


```{r}
shannon<-tm_shape(satract)+
  tm_polygons("var_land",style="quantile", n=5,legend.hist=T )+
  tm_shape(tractlines)+
  tm_lines(col="black")+
  tm_format("World",
            title="San Antonio Shannon Diversity of Land Area",
            legend.outside=TRUE )

shannon
```

## Exporting images for posters/presentations
I recommend using the `ggsave()` function to export images made by ggplot for use in other documents. For `tmap`, you can use `tmap_save()`

```{r,eval=FALSE}
#ggsave(m1, path = "", filename = "sa_pct_devel.png", units = "in", width=8, height = 10)
tmap_save(shannon,filename = "~/OneDrive - University of Texas at San Antonio/classes/gis_classwork/shannon_tract.png", width = 8, height = 10, units = "in")
```

And you can insert that image into a doc or presentation.


## Landsat imagery
Good simple description of landsat [here](https://gisgeography.com/landsat-8-bands-combinations/)

```{r}
library(raster)
library(rgdal)
library(rgeos)
library(RColorBrewer)


# turn off factors
options(stringsAsFactors = FALSE)
landsatstack<- list.files("~/OneDrive - University of Texas at San Antonio/classes/gis_classwork/LC08_L1TP_027040_20200311_20200325_01_T1/", 
  pattern=glob2rx("*B*.TIF$"), 
  full.names = T)

landsatstack
lsb2<-raster(landsatstack[2])

plot(lsb2,col = gray(0:100 / 100))  
  rm(lsb2)      ; gc()
```

## Make raster stack of bands 1 - 7
```{r}
lsstack<-stack(landsatstack[c(1,4:9)])

lsbrick<-brick(lsstack)
names(lsbrick)

names(lsbrick) <- gsub(pattern = "LC08_L1TP_027040_20200311_20200325_01_T1_",
                       replacement = "",
                       names(lsbrick))
names(lsbrick)
```

```{r}
ndvi<-(lsbrick[[5]] - lsbrick[[4]]) / (lsbrick[[5]] + lsbrick[[4]])

# plot(ndvi,
#      main = "NDVI around San Antonio",
#      axes = FALSE, box = FALSE)       
library(tmaptools)
tm_shape(ndvi)+
  tm_raster(style="quantile", n=5,legend.hist=T,palette =get_brewer_pal("RdYlGn", n = 5) )+
  tm_format("World",
            title="San Antonio Shannon Diversity of Land Area",
            legend.outside=TRUE )+
  tm_polygons(satract)


hist(ndvi)
```

```{r}
library(mapview)
mapview(nlcdreclass)
writeRaster(x= ndvi, 
    filename = "~/OneDrive - University of Texas at San Antonio/classes/gis_classwork/ndvi1.tif", format="GTiff", datatype="INT2S", overwrite=T)
```

### Landsat plots
```{r}
thermavg<-
plotRGB(lsbrick,
r = 6, g = 5, b = 2,
stretch = "lin",
axes = TRUE,
main = "RGB composite image\n showing Agricultural land - Landsat Bands 6, 5, 2")
box(col = "white")

plotRGB(lsbrick,
r = 5, g = 4, b = 3,
stretch = "lin",
axes = TRUE,
main = "Color infrared composite image\n Landsat Bands 5, 4, 3")
box(col = "white")

```
