---
title: "DEM 7093 GIS Summer 2018 - R mapping examples using American Community Survey - Change Mapping"
author: "Corey S. Sparks, PhD"
date: "June 6, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This example will use R to downloard American Coummunity Survey summary file tables using the `tidycensus` package. The goal of this example is to illustrate how to download data from the Census API using R and to illustrate how to create basic descriptive maps of attributes. 

The example will use data from San Antonio, Texas from the 2015 American Community Survey summary file.

###Get a Census developer API Key
Obtain one at http://api.census.gov/data/key_signup.html

###Save your API key to your working directory
use `census_api_key(key =  "yourkeyhere", install = T)`

one time to install your key for use in `tidycensus`



```{r, eval=FALSE, include=FALSE}
#install.packages("censusapi")
#don't use my key, put yours in here
#use install=T one time only
#library(censusapi)
census_api_key(key =  "f67be5a86e696eebf0c84a5a20d85f2c04f7e9ef")
```




###Look at available ACS variables
```{r}
library(tidycensus);  library(tidyverse); library(sf)
v15_Profile <- load_variables(2015, "acs5/profile", cache = TRUE) #demographic profile tables
v15_tables <- load_variables(2015 , "acs5", cache = TRUE) #all tables
#v10_sf1_tables <- load_variables(2010 , "sf1", cache = TRUE) #all tables for 2010 SF1
#v10_sf1_tables <- load_variables(2000 , "sf3", cache = TRUE) #all tables for 2000 SF#

View(v15_Profile)

#Search for variables by 
v15_Profile[grep(x = v15_Profile$label, "HighSchool"), c("name", "label")]
v15_Profile[grep(x = v15_Profile$label, "Built 2000 to 2009"), c("name", "label")]
```

##Extract from ACS summary file data profile variables from 2015 for Bexar County, TX Census Tracts

The data profile tables are very useful because they contain lots of pre-calculated variables.

```{r, results='hide'}
sa_acs<-get_acs(geography = "tract", state="TX", county = c("Bexar"), year = 2015,
                variables=c( "DP05_0001E", "DP03_0009P", "DP03_0062E", "DP03_0119PE",
                            "DP05_0001E","DP02_0009PE","DP02_0008PE", "DP02_0040E","DP02_0038E",
                            "DP02_0066PE","DP02_0067PE","DP02_0080PE","DP02_0092PE",
                            "DP03_0005PE","DP03_0028PE","DP03_0062E","DP03_0099PE","DP03_0101PE",
                            "DP03_0119PE","DP04_0046PE","DP05_0072PE","DP05_0073PE",
                            "DP05_0066PE", "DP05_0072PE", "DP02_0113PE") ,
                summary_var = "B01001_001",
                geometry = T, output = "wide")

#create a county FIPS code - 5 digit
sa_acs$county<-substr(sa_acs$GEOID, 1, 5)

#rename variables and filter missing cases
sa_acs2<-sa_acs%>%
  mutate(totpop= DP05_0001E,tpop_cv=(DP05_0001M/1.645)/DP05_0001E ,
         phsormore=DP02_0066PE,phsormore_cv=(DP02_0066PM/1.645)/DP02_0066PE, medhhinc=DP03_0062E,ppov=DP03_0119PE ) %>%
#  st_transform(crs = 102740)%>%
  filter(complete.cases(totpop, phsormore, medhhinc))

#(acs.sub$medhhinc_ME/1.645) / acs.sub$medhhinc) * 100

class(sa_acs2)
```

##Write data out to shapefile
```{r, message=FALSE, warning=FALSE, eval=F}
#change the directory
sf::st_write(sa_acs2,dsn="C:/Users/ozd504/Google Drive/classes/dem7093/GIS_class_2018/data",layer="sa_tract_dp", driver="ESRI Shapefile", delete_layer=T, update=T)

```

##mapping of errors in variables
Here I generate a quantile break for the coefficient of variation in census tract population estimates 
```{r, message=FALSE}

library(classInt)
library(dplyr)

cv_map<-sa_acs2 %>%
  mutate(cv_cut=cut(tpop_cv,breaks = quantile(tpop_cv, na.rm=T, p=seq(0,1,length.out = 6)),include.lowest = T),
         ed_cut=cut(phsormore_cv,breaks = quantile(phsormore_cv, na.rm=T, p=seq(0,1,length.out = 6)),include.lowest = T) )

library(ggsn)
p1<-ggplot(cv_map, aes(fill = cv_cut, color = cv_cut)) + 
  geom_sf() + 
  ggtitle("Coefficent of Variation in Population Estimate", 
          subtitle = "Bexar County Texas, 2015 ACS")+
    scale_fill_brewer(palette = "Blues") + 
  scale_color_brewer(palette = "Blues")+
  theme(axis.text.x = element_blank(), axis.text.y = element_blank())+
  north(cv_map)+
  scalebar(cv_map, dist = 5,  dd2km =T, model="GRS80", st.size = 2)
p1



p2<-ggplot(cv_map, aes(fill = ed_cut, color = ed_cut)) + 
  geom_sf() + 
  ggtitle("Coefficent of Variation in Estimate of % Greater than High School", 
          subtitle = "Bexar County Texas, 2015 ACS")+
  scale_fill_brewer(palette = "Blues") + 
  scale_color_brewer(palette = "Blues")+
    theme(axis.text.x = element_blank(), axis.text.y = element_blank())+
  north(cv_map)+
  scalebar(cv_map, dist = 5,  dd2km =T, model="GRS80", st.size = 2)
p2
  
```

##Change map between two time points
When we have data that are collected over time on the same geographies, we may be interested in whether the variable we're mapping has changed much over time. 

In the ACS, we can compare two estimates if the years used to produce the esimates do not overlap. For instance, we could compare the 2005-2009 estimates to the 2011-2015 estimates, but we could **not** compare the 2010-2014 to the 2011-2015, becuase they share years of data. See [this](https://www.census.gov/programs-surveys/acs/guidance/comparing-acs-data.html) for the official position on the subject.


##compare median household incomes
Here we take the median houshold income in tracts derived from the 2007-2011 ACS and compare it to the estimate from the 2011-2015 ACS
```{r}
#v11_Profile <- load_variables(2011 ,dataset =  "acs5") #demographic profile 

options(tigris_use_cache = TRUE)
#2009 ACS

acs_11<-get_acs(geography = "tract", state="TX", county = "029", year = 2011, variables="B19013_001",geometry = TRUE, output = "wide")

#2015 ACS
acs_15<-get_acs(geography = "tract", state="TX", county = "Bexar", year = 2015, variables="B19013_001",geometry = TRUE, output = "wide")

#create variables with nice names
acs_11<-acs_11%>%
  mutate(inc11=B19013_001E, inc11err=B19013_001M)%>%
  select(GEOID, NAME, inc11, inc11err)
#create variables with nice names
acs_15<-acs_15%>%
  mutate(inc15=B19013_001E, inc15err=B19013_001M)%>%
  select(GEOID, NAME, inc15, inc15err)

#merge the two years worth of data
mdat<-left_join(acs_11, as.data.frame(acs_15), by="GEOID", st_join=FALSE)

head(mdat)
```


Here I create a function that implements the [testing  procedure](https://www.psc.isr.umich.edu/dis/acs/handouts/Compass_Appendix.pdf) used by the Census for comparing estimates across year
```{r}
acstest<-function(names,geoid, est1, err1, est2, err2, alpha, yr1, yr2, span){
  se1<-err1/qnorm(.90)
  se2<-err2/qnorm(.90)
  yrs1<-seq(yr1, to=yr1-span)
  yrs2<-seq(yr2, to=yr2-span)

  C<-mean(yrs2%in%yrs1)
  diff<- (est1-est2)
  test<-(est1-est2) / (sqrt(1-C)*sqrt((se1^2+se2^2)))
  crit<-qnorm(1-alpha/2)
  pval<-2*pnorm(abs(test),lower.tail=F)
  result<-NULL
  result[pval > alpha]<-"insignificant change"
  result[pval < alpha & test > 0]<- "significant increase"
  result[pval < alpha & test < 0]<-"significant decrease" 
  
  data.frame(name=names,geoid=geoid, est1=est1, est2=est2, se1=se1, se2=se2,difference=diff, test=test, result=result, pval=pval)
}
```


Here I use the function I just made to do the comparisons
```{r}
diff0915<-acstest(names = mdat$NAME.x, geoid = mdat$GEOID, est1 = mdat$inc11*1.07, est2 = mdat$inc15, err1 = mdat$inc11err*1.12, err2=mdat$inc15err, alpha = .1, yr1 = 2011, yr2=2015, span = 5)

head(diff0915)

```

```{r, fig.height=10, fig.width=12}
options(scipen=999)
acs_merge<-left_join(sa_acs, diff0915, by=c("GEOID"="geoid"))
brks<-quantile(c(acs_11$inc11, acs_15$inc15),  probs = seq(0,1,.2), na.rm=T)
acs_11map<-acs_11%>%
  mutate(inc_q = cut(inc11,breaks = brks, include.lowest = T))

p1<-ggplot(acs_11map, aes( fill=inc_q, color=inc_q))+
  geom_sf() + 
  ggtitle("Median Household Income 2011 ", 
          subtitle = "Bexar County Texas")+
  scale_fill_brewer(palette = "Blues") + 
  scale_color_brewer(palette = "Blues")+
  #scale_fill_manual(values=myfil)+
    theme(axis.text.x = element_blank(), axis.text.y = element_blank())+
  north(acs_11map)+
  scalebar(acs_11map, dist = 5,  dd2km =T, model="GRS80", st.size = 2)

acs_15map<-acs_15%>%
  mutate(inc_q = cut(inc15,breaks = brks, include.lowest = T))

p2<-ggplot(acs_15map, aes( fill=inc_q, color=inc_q))+
  geom_sf() + 
  ggtitle("Median Household Income 2015 ", 
          subtitle = "Bexar County Texas")+
  scale_fill_brewer(palette = "Blues") + 
  scale_color_brewer(palette = "Blues")+
  #scale_fill_manual(values=myfil)+
    theme(axis.text.x = element_blank(), axis.text.y = element_blank())+
  north(acs_15map)+
  scalebar(acs_15map, dist = 5,  dd2km =T, model="GRS80", st.size = 2)


myfil<-c("grey", "red", "blue")
p3<-ggplot(acs_merge, aes( fill=result))+
  geom_sf() + 
  ggtitle("Comparison of Median Household Income 2011 to 2015", 
          subtitle = "Bexar County Texas")+
  #scale_fill_brewer(palette = "Accent") + 
  #scale_color_brewer(palette = "Accent")+
  scale_fill_manual(values=myfil)+
    theme(axis.text.x = element_blank(), axis.text.y = element_blank())+
  north(acs_merge)+
  scalebar(acs_merge, dist = 5,  dd2km =T, model="GRS80", st.size = 2)


library(gridExtra)

out<-grid.arrange(p1, p2, p3, nrow = 2)
out
```

