---
title: "DEM 7093 - Using R for Point analysis Part 1"
author: "Corey S. Sparks, Ph.D. - University of Texas at San Antonio"
date: "June 12, 2018"
output: html_document
---

In this example I will use QGIS geoprocessing scripts through the `RQGIS` library in R. I will use data from the 2005 Peru Demographic and Health Survey, where I construct Voronoi polygons from the primary sampling unit locations and generate estimates of educational attainment for women.

```{r}

library(sf)
library(RQGIS) #you need Qgis 2.18 installed, not 3!
library(mapview)
```

```{r}
peru_dhs_points<-st_read("P:/LAPIVpaper/PEGE5BFL", "PEGE52FL")
peru_dhs_points<-st_transform(peru_dhs_points, crs=24892)
#project the data into a meter based system
mapview(peru_dhs_points["DHSCLUST"])
```


##Set up QGIS environment
This lets R find where your QGIS binaries are located. `set_env()` should work without you specifying the path
```{r}
my_env<-set_env()

```

#Buffer analysis
First we find the name of the algorithm for point buffering
```{r}
find_algorithms(search_term = "buffer", qgis_env = my_env)

```
For this, we'll use the `qgis:fixeddistancebuffer` function, but we need to see what the arguments to the function are:

```{r}

get_usage(alg="qgis:fixeddistancebuffer", qgis_env = my_env, intern = F)
```
so we have 5 arguments, when we use the function, we need to specify all of these:

```{r}

params <- get_args_man(alg = "qgis:fixeddistancebuffer", qgis_env = my_env)
params

```

Here I do a 5km buffer around each PSU location.

```{r}
wd<-"C:/Users/ozd504/Google Drive/classes/dem7093/GIS_class_2018/data"
params$INPUT <-peru_dhs_points
params$DISTANCE<-5000 #5km around each point
params$OUTPUT<-file.path(wd, "peru_psu_buffer_5k.shp") # path to the output shapefile
```

now we have our parameters defined, we run the script:
```{r}
peru_buff <- run_qgis(alg = "qgis:fixeddistancebuffer",
                    params = params,
                    load_output = TRUE,
                    qgis_env = my_env)

mapview(peru_buff["DHSCLUST"], legend=T,map.types="OpenStreetMap")

```

5km buffers done! Now, If I only had some other point data, I could do some point in polygon operations!

##Vonoroi polygons
```{r}
find_algorithms(search_term = "Voronoi")
get_usage(alg="qgis:voronoipolygons", qgis_env = my_env, intern = F)
```
so we have 3 arguments, when we use the function,:

```{r}

params <- get_args_man(alg = "qgis:voronoipolygons", qgis_env = my_env)
params

```


```{r}
wd<-"C:/Users/ozd504/Google Drive/classes/dem7093/GIS_class_2018/data"
params$INPUT <-peru_dhs_points
params$OUTPUT<-file.path(wd, "peru_psu_von_poly.shp") # path to the output shapefile
```

now we have our parameters defined, we run the script:
```{r}
peru_von <- run_qgis(alg = "qgis:voronoipolygons",
                    params = params,
                    load_output = TRUE,
                    qgis_env = my_env)

mapview(peru_von["DHSCLUST"],map.types="OpenStreetMap")

```


The polygons obviously have issues, so we can clip them to the Peruvian border:

```{r}
find_algorithms(search_term = "clip")
get_usage(alg="qgis:clip", qgis_env = my_env, intern = F)
```
so we have 3 arguments, when we use the function,:

```{r}

params <- get_args_man(alg = "qgis:clip", qgis_env = my_env)
params

```


```{r}
peru_border<-st_read("P:/LAPIVpaper/PER_adm", "PER_adm0")
peru_border<-st_transform(peru_border,crs=24892)

wd<-"C:/Users/ozd504/Google Drive/classes/dem7093/GIS_class_2018/data"
params$INPUT <-peru_von
params$OVERLAY<-peru_border
params$OUTPUT<-file.path(wd, "peru_psu_von_poly_clip.shp") # path to the output shapefile
```

now we have our parameters defined, we run the script:
```{r}
peru_von_clip <- run_qgis(alg = "qgis:clip",
                    params = params,
                    load_output = TRUE,
                    qgis_env = my_env)

mapview(peru_von_clip["DHSCLUST"],map.types="OpenStreetMap")

```

##Map some estimates:

If we have the 2005 DHS data, we could generate estimates of some quality, say proportion of women with a secondary education or more:

```{r}
library(haven)
library(survey)
library(dplyr)

peru_data<-read_dta("P:/LAPIVpaper/DHS_IPVdata/PEIR51FL.DTA")

peru_data<-peru_data%>%
  mutate(eduprim=ifelse(v106==1,1,0),
  edusecplus=ifelse(v106!=9&v106>=2,1,0),
  pwt=v005/1000000)%>%
  select(v000, v021, v022,v024,pwt, eduprim, edusecplus )

head(peru_data)

library(lme4)
fit<-glmer(edusecplus~1+(1|v022/v021),data=peru_data, family = binomial, weights=pwt)

peru_data$edu_ests<-fitted(fit)
place_ests<-aggregate(edu_ests~v021, data=peru_data, FUN=mean)
head(place_ests)
```

Merge the estimates to the polygons
```{r}
peru_ests<-left_join(peru_von_clip, place_ests, by=c("DHSCLUST"= "v021"))


mapview(peru_ests["edu_ests"], legend=T,map.types="OpenStreetMap")
```