---
title: "DEM 7093 GIS Summer 2018 - R mapping examples using American Community Survey"
author: "Corey S. Sparks, PhD"
date: "June 4, 2018"
output:
  word_document: default
  html_document: default
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Objectives

This lesson has two objectives:

  1. Use the `tidycensus` package to download Census data
  2. Use `ggplot` and `mapview` functions to produce maps of these data


This example will use R to downloard American Coummunity Survey summary file tables using the `tidycensus` package. The goal of this example is to illustrate how to download data from the Census API using R and to illustrate how to create basic descriptive maps of attributes. 

The example will use data from San Antonio, Texas from the 2015 American Community Survey summary file.

###Get a Census developer API Key
Obtain one at http://api.census.gov/data/key_signup.html

###Save your API key to your working directory
use `census_api_key(key =  "yourkeyhere", install = T)`

one time to install your key for use in `tidycensus`



```{r, eval=FALSE, include=FALSE}
#install.packages("censusapi")
#don't use my key, put yours in here
#use install=T one time only
#library(censusapi)
census_api_key(key =  "f67be5a86e696eebf0c84a5a20d85f2c04f7e9ef")
```

\newpage


### Look at available ACS variables
We may not know the variable name we need for a particular project, but there are functions in the `tidycensus` library to help find variables. For example, if we are interested in variables from the ACS [data profile](https://www.census.gov/acs/www/data/data-tables-and-tools/data-profiles/2016/) tables, we can load all available variables then use R to search for what we need.


```{r, message=FALSE}
library(tidycensus);  library(tidyverse); library(sf)
v15_Profile <- load_variables(2015 , "acs5/profile", cache = TRUE) #demographic profile tables

View(v15_Profile)

#Search for variables by keywords in the label 
v15_Profile[grep(x = v15_Profile$label, "POVERTY"), c("name", "label")]
v15_Profile[grep(x = v15_Profile$label, "Built 2000 to 2009"), c("name", "label")]
```

\newpage
## Extract from ACS summary file data profile variables from 2015 for Bexar County, TX Census Tracts

The data profile tables are very useful because they contain lots of pre-calculated variables.

Here is a query where we extract several variables from tehe 2015 ACS for Bexar County, Texas. We can also get the spatial data by requesting `geometry=TRUE`.  Useing `output="wide"` will put each variable in a column of the data set, with each row being a census tract. 

```{r, results='hide'}
sa_acs<-get_acs(geography = "tract", state="TX", county = c("Bexar"), year = 2015,
                variables=c( "DP05_0001E", "DP03_0009P", "DP03_0062E", "DP03_0119PE",
                           "DP05_0001E","DP02_0009PE","DP02_0008PE","DP02_0040E","DP02_0038E",
                            "DP02_0066PE","DP02_0067PE","DP02_0080PE","DP02_0092PE",
                        "DP03_0005PE","DP03_0028PE","DP03_0062E","DP03_0099PE","DP03_0101PE",
                            "DP03_0119PE","DP04_0046PE","DP05_0072PE","DP05_0073PE",
                            "DP05_0066PE", "DP05_0072PE", "DP02_0113PE") ,
                geometry = T, output = "wide")

#create a county FIPS code - 5 digit
sa_acs$county<-substr(sa_acs$GEOID, 1, 5)

#rename variables and filter missing cases
sa_acs2<-sa_acs%>%
  mutate(totpop= DP05_0001E, fertrate = DP02_0040E,pwhite=DP05_0072PE, 
         pblack=DP05_0073PE , phisp=DP05_0066PE, pfemhh=DP02_0008PE,
         phsormore=DP02_0066PE,punemp=DP03_0009PE, medhhinc=DP03_0062E,
         ppov=DP03_0119PE, pforn=DP02_0092PE,plep=DP02_0113PE) %>%
#  st_transform(crs = 102740)%>%
  na.omit()

```

## Write data out to shapefile
```{r, message=FALSE, warning=FALSE, eval=F}
#change the directory
sf::st_write(sa_acs2,dsn="C:/Users/ozd504/Google Drive/classes/dem7093/GIS_class_2018/data",layer="sa_tract_dp", driver="ESRI Shapefile", delete_layer=T, update=T)

```

\newpage

## Some basic mapping of variables
Here I generate a quantile break for % black in census tracts and compare it to a Jenks break
```{r, message=FALSE}
#install.packages("classInt")
#install.packages("patchwork")

library(classInt)
library(patchwork)

#install.packages("dplyr")
library(dplyr)

pblack_map<-sa_acs2 %>%
  mutate(cblack=cut(pblack,breaks = quantile(pblack, na.rm=T, p=seq(0,1,length.out = 6)),include.lowest = T),
           jblack = cut(pblack,breaks=data.frame(classIntervals(var=sa_acs2$pblack, n=5, style="jenks")[2])[,1], include.lowest = T))
library(ggsn)
p1<-ggplot(pblack_map, aes(fill = cblack, color = cblack)) + 
  geom_sf() + 
  ggtitle("Proportion African American", 
          subtitle = "Bexar County Texas, 2015 - Quantile Breaks")+
    scale_fill_brewer(palette = "Blues") + 
  scale_color_brewer(palette = "Blues")+
  theme(axis.text.x = element_blank(), axis.text.y = element_blank())+
  north(pblack_map)+
  scalebar(pblack_map, dist = 5,  dd2km =T, model="GRS80", st.size = 2)
p1



p2<-ggplot(pblack_map, aes(fill = jblack, color = jblack)) + 
  geom_sf() + 
  ggtitle("Proportion African American", 
          subtitle = "Bexar County Texas, 2015 - Jenks Breaks")+
  scale_fill_brewer(palette = "Blues") + 
  scale_color_brewer(palette = "Blues")+
    theme(axis.text.x = element_blank(), axis.text.y = element_blank())+
  north(pblack_map)+
  scalebar(pblack_map, dist = 5,  dd2km =T, model="GRS80", st.size = 2)
p2
  
p1/ p2

```

\newpage

## Interactive map with mapview

```{r}
library(mapview)
library(RColorBrewer)
pal <- colorRampPalette(brewer.pal(6, "Blues")) #set colors
mapview(pblack_map["jblack"], col.regions=pal, legend=T,map.types="OpenStreetMap", layer.name="% African American")


```



