---
title: "DEM 7093 - Using R for Point analysis Part 2 - Derived Variables"
author: "Corey S. Sparks, Ph.D. - University of Texas at San Antonio"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
   html_document:
    df_print: paged
    fig_height: 7
    fig_width: 7
    toc: yes
    toc_float: yes
---

In this example I will use QGIS geoprocessing scripts through the `qgisprocess` library in R. I will use data from the 2005 Peru Demographic and Health Survey, and data from the Peruvian government on the locations of [secondary schools](http://sigmed.minedu.gob.pe/mapaeducativo/). 


We use buffers from each DHS primary sampling unit location and point in polygon operations to measure whether a community had a secondary school within 5km. 

Then, we use a hierarchical model to test whether a woman's educational attainment is related to physical access to secondary schooling.

This is an example of a derived variable that cannot be obtained without the use of the GIS. 

```{r}

library(sf)
library(qgisprocess) 
library(tmap)
library(dplyr)

qgis_configure()
```
## Read in the DHS sampling unit locations
```{r}
peru_dhs_points<-st_read("~/OneDrive - University of Texas at San Antonio/projects/LAPIVpaper/PEGE52FL/PEGE52FL.shp", "PEGE52FL")

peru_dhs_points<-st_transform(peru_dhs_points, crs=24892)

#project the data into a meter based system
peru_dhs_points<-peru_dhs_points%>%
  filter(LATNUM<0)
```




## Point in Polygon analysis
So, right now we only have points (locations of sampling units), so we need a buffer around each point in order to do our assessment of whether a school is within 5km of each community. To do this we will do a fixed distance buffer of 5km around each sampling location. 

## Buffer analysis
Here I do a 5km buffer around each PSU location.
```{r}
peru_buff<-st_buffer(peru_dhs_points, dist = 5000)

```


now we have our parameters defined, we run the script:
```{r}
tmap_mode("view")

tm_basemap("OpenStreetMap.Mapnik")+
  tm_shape(peru_buff)+
  tm_polygons("DHSCLUST")+
  tm_shape(peru_dhs_points)+
  tm_dots()
  

```

5km buffers done! Now, we need to do our point in polygon operation. So I read in the Peruvian school data. These are a csv file, but have lat/long specified, so I can read in the text and make a simple feature layer from it. You can't have missing information in the coordinate data to do this.

```{r}
library(dplyr)
schools<-read.csv("~/OneDrive - University of Texas at San Antonio//classes/dem7093/dem7093_21//data/perusecondaryschools.csv", header=T)
schools<-schools%>%
  filter(complete.cases(Latitud, Longitud))

persch<-st_as_sf(schools, coords=c("Longitud", "Latitud"), crs=4326,agr="constant")
persch<-st_transform(persch,crs=24892 )


tm_basemap("OpenStreetMap.Mapnik")+
  #tm_shape(peru_buff)+
  #tm_polygons("DHSCLUST")+
  tm_shape(persch)+
  tm_dots("Departamento")


```

## points in polygons
See what the right script is, and what arguments it needs:

```{r}
peru_buff$nsch <- lengths(st_intersects(peru_buff, persch))
hist(peru_buff$nsch)
peru_buff$closeschool<-ifelse(peru_buff$nsch>0,1,0)
summary(peru_buff$closeschool)
```



So we see that 93.2% of communities have a secondary school within 5km, that's great, but what about the communities that don't?


## Merge the new location data to our survey

Here we load the 2005 DHS data, and do some recodes for educational attainment, age, and get the survey design variables. 

```{r}
library(haven)
library(survey)
library(dplyr)

peru_data<-read_dta("~/OneDrive - University of Texas at San Antonio/projects/LAPIVpaper/DHS_IPVdata/PEIR51FL.DTA")

peru_data2<-peru_data%>%
  mutate(eduprim=ifelse(v106==1,1,0),
  edusecplus=ifelse(v106!=9&v106>=2,1,0),
  pwt=v005/1000000, 
  modcontra = ifelse(v313==3, 1, 0),
  purchdes = car::Recode(v743b, recodes = "1:3=1;4:6=0;0=NA;else=NA"),
  age=v012,
  knowhiv = ifelse(v751==1, 0, 0),
  ipv.viol=ifelse(d105a==1|d105a==2|d105b==1|d105b==2| d105c==1|d105c==2 | d105d==1|d105d==2| d105e==1|d105e==2| d105g==1|d105g==2,1,0))%>%
  dplyr::select(v000, v021, v022,v024,age, pwt, eduprim, edusecplus, modcontra, purchdes, knowhiv, ipv.viol )

head(peru_data2)
library(survey)

```

Now we merge the survey to the spatial data:
```{r}

peru_merge<-left_join(peru_data2, peru_buff,by=c("v021"="DHSCLUST"))

```

In order to test for the effect of school access on women's educational attainment, we use a binomial generalized linear model. 

```{r}
des<-svydesign(ids=~v021, strata=~v022, weights=~pwt, data=peru_merge, nest = T)

fit<-svyglm(edusecplus~closeschool+age+I(age^2), family = binomial, design=des)

summary(fit)

```
So, we see in this case that if women lived within 5km of a secondary school, they are much more likely to have a secondary education or more, compared to woman living in a community without a school within 5km. 
