---
title: "DEM 7093 - Using R for Point analysis Part 1"
author: "Corey S. Sparks, Ph.D. - University of Texas at San Antonio"
date: "June 12, 2018"
output: html_document
---

In this example I will use QGIS geoprocessing scripts through the `RQGIS` library in R. I will use data from the 2005 Peru Demographic and Health Survey, where I construct Voronoi polygons from the primary sampling unit locations and generate estimates of educational attainment for women.

```{r}

library(sf)
library(RQGIS3) 
library(dplyr)
library(mapview)
```

```{r}
peru_dhs_points<-st_read("C:/Users/ozd504/Google Drive/classes/dem7093/dem7093_20/data/PEGE5BFL", "PEGE52FL")
peru_dhs_points<-st_transform(peru_dhs_points, crs=24892)
#project the data into a meter based system
mapview(peru_dhs_points["DHSCLUST"])
```


##Set up QGIS environment
This lets R find where your QGIS binaries are located. `set_env()` should work without you specifying the path
```{r}
my_env<-set_env()
open_app()
```

#Buffer analysis
First we find the name of the algorithm for point buffering
```{r}
find_algorithms(search_term = "buffer", qgis_env = my_env)

```
For this, we'll use the `native:buffer` function, but we need to see what the arguments to the function are:

```{r, results='hide'}

#get_usage(alg="native:buffer")
```
so we have 5 arguments, when we use the function, we need to specify all of these:

```{r}

params <- get_args_man(alg = "native:buffer")

```

Here I do a 5km buffer around each PSU location.

```{r}
wd<-"C:/Users/ozd504/OneDrive - University of Texas at San Antonio/gis_classwork/"
params$INPUT <-peru_dhs_points
params$DISTANCE<-5000 #5km around each point
params$OUTPUT<-file.path(wd, "peru_psu_buffer_5k.shp") # path to the output shapefile
```

now we have our parameters defined, we run the script:
```{r}
peru_buff <- run_qgis(alg = "native:buffer",
                    params = params,
                    load_output = TRUE,
                    qgis_env = my_env)
st_crs(peru_buff)<-24892
mapview(peru_buff["DHSCLUST"], legend=T,map.types="OpenStreetMap")

```

## buffer using R
```{r}
buf2<-st_buffer(peru_dhs_points, dist = 5000)
mapview(buf2["DHSCLUST"])
```
5km buffers done! Now, If I only had some other point data, I could do some point in polygon operations!

##Vonoroi polygons
```{r}
find_algorithms(search_term = "Voronoi")
#get_usage(alg="qgis:voronoipolygons", qgis_env = my_env, intern = F)
```
so we have 3 arguments, when we use the function,:

```{r}

params <- get_args_man(alg = "qgis:voronoipolygons", qgis_env = my_env)


```


```{r}
params$INPUT <-peru_dhs_points
params$OUTPUT<-file.path(wd, "peru_psu_von_poly.shp") # path to the output shapefile
```

now we have our parameters defined, we run the script:
```{r}
peru_von <- run_qgis(alg = "qgis:voronoipolygons",
                    params = params,
                    load_output = TRUE,
                    qgis_env = my_env)

st_crs(peru_von)<-24892
mapview(peru_von["DHSCLUST"],map.types="OpenStreetMap")

```

### I can't figure out how to use R for this

The polygons obviously have issues, so we can clip them to the Peruvian border:

```{r}
find_algorithms(search_term = "clip")
#get_usage(alg="native:clip")
```

so we have 3 arguments, when we use the function,:

```{r}

params <- get_args_man(alg = "native:clip")


```


```{r}
peru_border<-st_read("C:/Users/ozd504/Google Drive/classes/dem7093/dem7093_20/data/PER_adm", "PER_adm0")
peru_border<-st_transform(peru_border,crs=24892)

params$INPUT <-peru_von
params$OVERLAY<-peru_border
params$OUTPUT<-file.path(wd, "peru_psu_von_poly_clip.shp") # path to the output shapefile
```

now we have our parameters defined, we run the script:
```{r}
peru_von_clip <- run_qgis(alg = "native:clip",
                    params = params,
                    load_output = TRUE,
                    qgis_env = my_env)

mapview(peru_von_clip["DHSCLUST"],map.types="OpenStreetMap")


```

### Using R
You can use st_clip to clip observations to a polygon
```{r}
pclip<-st_intersection(peru_von, peru_border)
# pclip2<-peru_von%>%
#   filter(st_contains( peru_border, ., sparse = F))

plot(pclip["DHSCLUST"])

```

##points in polygons
See what the right script is, and what arguments it needs:

```{r}
find_algorithms(search_term = "in polygon")
#get_usage(alg="qgis:countpointsinpolygon", qgis_env = my_env, intern = F)
```

```{r}

params <- get_args_man(alg = "qgis:countpointsinpolygon", qgis_env = my_env)


```

5km buffers done! Now, we need to do our point in polygon operation. So I read in the Peruvian school data. These are a csv file, but have lat/long specified, so I can read in the text and make a simple feature layer from it. You can't have missing information in the coordinate data to do this.

```{r}

schools<-read.csv("C:/Users/ozd504/Google Drive/classes/dem7093/dem7093_20/data/perusecondaryschools.csv", header=T)
schools<-schools%>%
  filter(complete.cases(Latitud, Longitud))

persch<-st_as_sf(schools, coords=c("Longitud", "Latitud"), crs=4326,agr="constant")
persch<-st_transform(persch,crs=24892 )

mapview(persch["Departamento"])
```

##make the parameter block
```{r}
params$POLYGONS <-peru_buff
params$POINTS<-persch
params$OUTPUT<-file.path(wd, "peru_schoolsinbuffer.shp") # path to the output shapefile
```

now we have our parameters defined, we run the script:
```{r}
peru_pip <- run_qgis(alg = "qgis:countpointsinpolygon",
                    params = params,
                    load_output = TRUE,
                    qgis_env = my_env)

mapview(peru_pip["NUMPOINTS"],map.types="OpenStreetMap", legend=T)

#recode the count of schools to be binary (0 or more than 0 schools)
peru_pip$closeschool<-ifelse(peru_pip$NUMPOINTS>0,1,0)
summary(peru_pip$closeschool)
```

So we see that 92.8% of communities have a secondary school within 5km, that's great, but what about the communities that don't?


## using R

```{r}
sfpip<-lengths(st_intersects( peru_buff,persch))
peru_buff$nschools<-sfpip
hist(peru_buff$nschools)
```

## use these in an analysis

Here we load the 2005 DHS data, and do some recodes for educational attainment, age, and get the survey design variables. 

```{r}
library(haven)
library(survey)
library(dplyr)

peru_data<-read_dta("C:/Users/ozd504/Google Drive/classes/dem7093/dem7093_20//data/PEIR51FL.DTA")

peru_data<-peru_data%>%
  mutate(eduprim=ifelse(v106==1,1,0),
  edusecplus=ifelse(v106!=9&v106>=2,1,0),
  pwt=v005/1000000, 
  age=v012)%>%
  select(v000, v021, v022,v024,age, pwt, eduprim, edusecplus )

head(peru_data)

peru_data$one<-1
des<-svydesign(ids=~v021, strata=~v022, weights=~pwt, data=peru_data, nest = T)
counts<-svyby(~edusecplus+one, ~v021, des, FUN = svytotal)
head(counts)
```

Now we merge the survey to the spatial data:
```{r}

peru_merge<-left_join(peru_data,peru_pip,by=c("v021"="DHSCLUST"))

```

In order to test for the effect of school access on women's educational attainment, we use a binomial generalized linear mixed model. We control for the survey design using a nested random effect specification (PSUs within strata).

```{r}
des<-svydesign(ids=~v021, strata=~v022, weights=~pwt, data=peru_merge, nest = T)

fit<-svyglm(edusecplus~closeschool+scale(age), family = binomial, design=des)

summary(fit)

```
So, we see in this case that if women lived within 5km of a secondary school, they are much more likely to have a secondary education or more, compared to woman living in a community without a school within 5km. 

