---
title: "DEM 7093 - Using R for Point analysis Part 2 - Derived Variables"
author: "Corey S. Sparks, Ph.D. - University of Texas at San Antonio"
date: "June 13, 2018"
always_allow_html: yes
output:
  pdf_document: default
  html_document: default
---

In this example I will use QGIS geoprocessing scripts through the `RQGIS` library in R. I will use data from the 2005 Peru Demographic and Health Survey, and data from the Peruvian government on the locations of [secondary schools](http://sigmed.minedu.gob.pe/mapaeducativo/). 


We use buffers from each DHS primary sampling unit location and point in polygon operations to measure whether a community had a secondary school within 5km. 

Then, we use a hierarchical model to test whether a woman's educational attainment is related to physical access to secondary schooling.

This is an example of a derived variable that cannot be obtained without the use of the GIS. 

```{r}

library(sf)
library(RQGIS) #you need Qgis 2.18 installed, not 3!
library(mapview)
```
##Read in the DHS sampling unit locations
```{r}
peru_dhs_points<-st_read("P:/LAPIVpaper/PEGE5BFL", "PEGE52FL")
peru_dhs_points<-st_transform(peru_dhs_points, crs=24892)
#project the data into a meter based system
mapview(peru_dhs_points["DHSCLUST"])
```


##Set up QGIS environment
This lets R find where your QGIS binaries are located. `set_env()` should work without you specifying the path, but in my case I had to tell R where Qgis lives.

```{r}
my_env<-set_env()

```

##Point in Polygon analysis
So, right now we only have points (locations of sampling units), so we need a buffer around each point in order to do our assessment of whether a school is within 5km of each community. To do this we will do a fixed distance buffer of 5km around each sampling location. 

##Buffer analysis
First we find the name of the algorithm for point buffering
```{r}
find_algorithms(search_term = "buffer", qgis_env = my_env)

```
For this, we'll use the `qgis:fixeddistancebuffer` function, but we need to see what the arguments to the function are:

```{r}
get_usage(alg="qgis:fixeddistancebuffer", qgis_env = my_env, intern = F)
```
so we have 5 arguments, when we use the function, we need to specify *some* of these:

```{r}

params <- get_args_man(alg = "qgis:fixeddistancebuffer", qgis_env = my_env)
params

```

Here I do a 5km buffer around each PSU location.

```{r}
wd<-"C:/Users/ozd504/Google Drive/classes/dem7093/GIS_class_2018/data"
params$INPUT <-peru_dhs_points
params$DISTANCE<-5000 #5km around each point
params$OUTPUT<-file.path(wd, "peru_psu_buffer_5k.shp") # path to the output shapefile
```

now we have our parameters defined, we run the script:
```{r}
peru_buff <- run_qgis(alg = "qgis:fixeddistancebuffer",
                    params = params,
                    load_output = TRUE,
                    qgis_env = my_env)

mapview(peru_buff["DHSCLUST"], legend=T,map.types="OpenStreetMap")

```

5km buffers done! Now, we need to do our point in polygon operation. So I read in the Peruvian school data. These are a csv file, but have lat/long specified, so I can read in the text and make a simple feature layer from it. You can't have missing information in the coordinate data to do this.

```{r}
library(dplyr)
schools<-read.csv("C:/Users/ozd504/Google Drive/classes/dem7093/GIS_class_2018/data/bars.csv", header=T)

persch<-st_as_sf(schools, coords=c("Longitude", "Latitude"), crs=4269,agr="constant")
persch<-st_transform(persch,crs=102009 )

mapview(persch["Source"])
```

##points in polygons
See what the right script is, and what arguments it needs:

```{r}
find_algorithms(search_term = "in polygon")
get_usage(alg="qgis:countpointsinpolygon", qgis_env = my_env, intern = F)
```

```{r}

params <- get_args_man(alg = "qgis:countpointsinpolygon", qgis_env = my_env)
params

```

##make the parameter block
```{r}
wd<-"C:/Users/ozd504/Google Drive/classes/dem7093/GIS_class_2018/data"
params$POLYGONS <-peru_buff
params$POINTS<-persch
params$OUTPUT<-file.path(wd, "peru_schoolsinbuffer.shp") # path to the output shapefile
```

now we have our parameters defined, we run the script:
```{r}
peru_pip <- run_qgis(alg = "qgis:countpointsinpolygon",
                    params = params,
                    load_output = TRUE,
                    qgis_env = my_env)

mapview(peru_pip["NUMPOINTS"],map.types="OpenStreetMap", legend=T)

#recode the count of schools to be binary (0 or more than 0 schools)
peru_pip$closeschool<-ifelse(peru_pip$NUMPOINTS>0,1,0)
summary(peru_pip$closeschool)
```

So we see that 92.8% of communities have a secondary school within 5km, that's great, but what about the communities that don't?


##Merge the new location data to our survey

Here we load the 2005 DHS data, and do some recodes for educational attainment, age, and get the survey design variables. 

```{r}
library(haven)
library(survey)
library(dplyr)

peru_data<-read_dta("P:/LAPIVpaper/DHS_IPVdata/PEIR51FL.DTA")

peru_data<-peru_data%>%
  mutate(eduprim=ifelse(v106==1,1,0),
  edusecplus=ifelse(v106!=9&v106>=2,1,0),
  pwt=v005/1000000, 
  age=v012)%>%
  select(v000, v021, v022,v024,age, pwt, eduprim, edusecplus )

head(peru_data)

```

Now we merge the survey to the spatial data:
```{r}

peru_merge<-left_join(peru_data, peru_pip,by=c("v021"="DHSCLUST"))

```

In order to test for the effect of school access on women's educational attainment, we use a binomial generalized linear mixed model. We control for the survey design using a nested random effect specification (PSUs within strata).

```{r}
library(lme4)
fit<-glmer(edusecplus~closeschool+scale(age)+(1|v022/v021),data=peru_merge, family = binomial, weights=pwt)

summary(fit)

```
So, we see in this case that if women lived within 5km of a secondary school, they are `r 100*round( 1- exp(fixef(fit)[3]),4)` % more likely to have a secondary education or more, compared to woman living in a community without a school within 5km. 
